---
title: "EVs and local air quality"
author: "Erica Bishop"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(here)
library(janitor)
library(lubridate)
library(datapasta)

```


```{r}
#set up data directory
datadir <- ("/Users/ericabishop/Documents/MEDS-fall-classwork/EDS222-stats/final_project/data")


#read in EV data and format dates
co_evs <- read_csv(file.path(datadir, "co_ev_registrations_public.csv")) |> 
  janitor::clean_names() 

#format dates
co_evs <- co_evs |> 
  mutate(registration_valid_date = lubridate::mdy(registration_valid_date),
         registration_expiration_date = lubridate::mdy(registration_expiration_date))

```
EV data: 
- zip codes are a hot mess (multiple different formats / lengths?)
- over 1300 unique zips (should be no more than 500 or so)
- date range: January 2010 - July 2022

Air Quality data from the EPA's Air Data: https://www.epa.gov/outdoor-air-quality-data
- from https://aqs.epa.gov/aqsweb/airdata/download_files.html
- Daily AQI, Daily Ozone, and Annual AQI data by county


```{r}

#read in raw daily AQI data

daily_aqi_raw_10 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2010.csv"))
daily_aqi_raw_11 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2011.csv"))
daily_aqi_raw_12 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2012.csv"))
daily_aqi_raw_13 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2013.csv"))
daily_aqi_raw_14 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2014.csv"))
daily_aqi_raw_15 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2015.csv"))
daily_aqi_raw_16 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2016.csv"))
daily_aqi_raw_17 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2017.csv"))
daily_aqi_raw_18 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2018.csv"))
daily_aqi_raw_19 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2019.csv"))
daily_aqi_raw_20 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2020.csv"))
daily_aqi_raw_21 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2021.csv"))
daily_aqi_raw_22 <- read_csv(file.path(datadir, "daily_county_aqi", "daily_aqi_by_county_2022.csv"))

```

```{r}
#Function to filter dfs to just counties in Colorado
filter_CO <- function(df) {
  df_filtered <- df |> 
    filter(`State Name` == "Colorado") |> 
    clean_names()
  return(df_filtered)
}

#run function on data frames:
daqi_10_CO <- filter_CO(daily_aqi_raw_10)
daqi_11_CO <- filter_CO(daily_aqi_raw_11)
daqi_12_CO <- filter_CO(daily_aqi_raw_12)
daqi_13_CO <- filter_CO(daily_aqi_raw_13)
daqi_14_CO <- filter_CO(daily_aqi_raw_14)
daqi_15_CO <- filter_CO(daily_aqi_raw_15)
daqi_16_CO <- filter_CO(daily_aqi_raw_16)
daqi_17_CO <- filter_CO(daily_aqi_raw_17)
daqi_18_CO <- filter_CO(daily_aqi_raw_18)
daqi_19_CO <- filter_CO(daily_aqi_raw_19)
daqi_20_CO <- filter_CO(daily_aqi_raw_20)
daqi_21_CO <- filter_CO(daily_aqi_raw_21)
daqi_22_CO <- filter_CO(daily_aqi_raw_22)
  
#join using rbind
daqi_CO <- rbind(daqi_10_CO, daqi_11_CO, daqi_12_CO, daqi_13_CO, daqi_14_CO, daqi_15_CO, daqi_16_CO, daqi_17_CO, daqi_18_CO, daqi_19_CO, daqi_20_CO, daqi_21_CO, daqi_22_CO)

```

```{r}

#read in daily ozone data
daily_ozone_2010 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2010.csv"))
daily_ozone_2011 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2011.csv"))
daily_ozone_2012 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2012.csv"))
daily_ozone_2013 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2013.csv"))
daily_ozone_2014 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2014.csv"))
daily_ozone_2015 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2015.csv"))
daily_ozone_2016 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2016.csv"))
daily_ozone_2017 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2017.csv"))
daily_ozone_2018 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2018.csv"))
daily_ozone_2019 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2019.csv"))
daily_ozone_2020 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2020.csv"))
daily_ozone_2021 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2021.csv"))
daily_ozone_2022 <- read_csv(file.path(datadir, "daily_summary_ozone", "daily_44201_2022.csv"))

#filter to just CO
dozo_10_CO <- filter_CO(daily_ozone_2010)
dozo_11_CO <- filter_CO(daily_ozone_2011)
dozo_12_CO <- filter_CO(daily_ozone_2012)
dozo_13_CO <- filter_CO(daily_ozone_2013)
dozo_14_CO <- filter_CO(daily_ozone_2014)
dozo_15_CO <- filter_CO(daily_ozone_2015)
dozo_16_CO <- filter_CO(daily_ozone_2016)
dozo_17_CO <- filter_CO(daily_ozone_2017)
dozo_18_CO <- filter_CO(daily_ozone_2018)
dozo_19_CO <- filter_CO(daily_ozone_2019)
dozo_20_CO <- filter_CO(daily_ozone_2020)
dozo_21_CO <- filter_CO(daily_ozone_2021)
dozo_22_CO <- filter_CO(daily_ozone_2022)

#combine into one df with rbind
dozo_CO <- rbind(dozo_10_CO, dozo_11_CO, dozo_12_CO, dozo_13_CO, dozo_14_CO, dozo_15_CO, dozo_16_CO, dozo_17_CO, dozo_18_CO, dozo_19_CO, dozo_20_CO, dozo_21_CO)



```


### Part 1: Look for a model - is there a relationship between any of the variables?

- find correlation
- find covariance
- scatter plot

### Part 2: Build a model


### Part 3: Test Hypothesis

Null Hypothesis: beta_1 (slope) = 0 (no relationship between ozone and # of EVs)
Alternate Hypothesis: beta_1 != 0 (or is negative?)





